---
title: "Basics of Statistical Visualization with ggplot2"
author: "Alvaro Fuentes"
date: "2016"
output: html_document
---

# Introduction

```{r, echo=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw())

# Loading the data
path_to_data <- 'https://raw.githubusercontent.com/quantcompany/Springboard_FDS/master/titanic_original.csv'
titanic <- read.csv(path_to_data, stringsAsFactors = FALSE, na.strings = '')

```


#Univariate Plots: Quantitative Variables

## Histograms

Histograms are plots used to explore how quantitative variables are distributed. It is an estimate of the probability distribution of a continuous variable (quantitative variable). To construct a histogram, the first step is to "bin" the range of values—that is, divide the entire range of values into a series of intervals—and then count how many values fall into each interval. The bins are usually specified as consecutive, non-overlapping intervals of a variable. The bins (intervals) must be adjacent, and are usually equal size.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(titanic, aes(x=age)) +
    geom_histogram(color = 'black', fill='#DDDDDD', binwidth = 2) +
    ggtitle('Distribution of Ages in the Titanic')


ggplot(titanic, aes(x=age, fill=survived2)) +
    geom_histogram(color = 'black', binwidth = 2) +
    ggtitle('Distribution of Ages in the Titanic')


ggplot(titanic, aes(x = age)) +
    geom_histogram(color = 'black', fill = 'lightblue') +
    geom_text(stat = 'bin', vjust=-0.25, 
              aes(y = ..count.., label = ..count..)) +
    ggtitle('Distribution of Ages in the Titanic')
```

## Density Plots

density plots are usually a much more effective way to view the distribution of a variable

```{r, echo=FALSE, warning=FALSE}
ggplot(titanic, aes(x = age)) +
    geom_density(color = 'black', fill = 'lightblue', alpha = 0.95) +
    ggtitle('Distribution of Ages in the Titanic')
```

## Histogams and Density Plots

We can have them both, why not?
All we need to do is add both layers to the plot, however we need to add `y = ..density..` in the `aes` function so the height of the bars are not the counts but the density: estimation of the probability of the variable taking the values in the x axis.

```{r, echo=FALSE, warning=FALSE}
ggplot(titanic, aes(x = age, y = ..density..)) + 
    geom_histogram(color = 'black', fill = 'lightblue')  + 
    geom_density(color = 'black', fill = 'lightblue', alpha = 0.4) +
    ggtitle('Distribution of Ages in the Titanic')
```

#Univariate Plots: Qualitative Variables

## Bar charts


```{r, echo=FALSE, warning=FALSE}
titanic$sex <- factor(titanic$sex)

ggplot(titanic[!is.na(titanic$sex),], aes(x = sex)) +
    geom_bar() +
    geom_text(stat = 'count', aes(y = (..count.. + 30), label = ..count..)) +
    ggtitle('Genders of Titanic Passangers')

```

## Proportions

Don't use pie charts. 
```{r, warning=FALSE}

ggplot(titanic[!is.na(titanic$sex),], aes(x = 'gender', fill = sex)) +
    geom_bar(position = 'fill') + xlab('') + ylab('') +
    geom_text(stat = 'count',
              position = 'fill',
              hjust = c(2.2,4), size = 5,
              aes(ymax = ..count..,
                  label = paste0(round(100*..count../sum(..count..),1),'%'))
              ) + 
    coord_flip()

```

# Bivariate plots: Quantitative vs. Quantitative

```{r, echo=FALSE, warning=FALSE}

ggplot(titanic, aes(x = age, y = fare)) +
    geom_point()

```

# Bivariate plots: Qualitative vs. Qualitative

```{r, echo=FALSE, warning=FALSE}
titanic$survived2 <- factor(titanic$survived, levels = c(0,1), labels = c('No','Yes'))

ggplot(titanic[!is.na(titanic$survived2),], aes(x = survived2, fill = sex)) +
    geom_bar(position = 'stack') +
    geom_text(stat = 'count', position = 'stack', vjust = -0.25,
              aes(y = ..count.., label = ..count..))

ggplot(titanic[!is.na(titanic$survived2),], aes(x = survived2, fill = sex)) +
    geom_bar(position = 'dodge') +
    geom_text(stat = 'count',  vjust = -0.25, 
              position = position_dodge(width = 1),
              aes(y = ..count.., label = ..count..))

ggplot(titanic[!is.na(titanic$survived2),], 
            aes(x = survived2, fill = sex)) +
    geom_bar(position = 'fill') + xlab('Survived') + ylab('') +
    annotate('text', x = c(1,2,1,2), y = c(0.08,0.4,0.5,0.8), size = 5,
             label = c('15.7%','68.7%','84.3%','32.2%')) + 
    coord_flip()
```


# Bivariate plots: Quantitative vs. Qualitative

```{r, echo=FALSE, warning=FALSE}

ggplot(titanic[!is.na(titanic$survived2),], 
       aes(x = factor(pclass), y = age)) +
    geom_boxplot()

ggplot(titanic[!is.na(titanic$survived2),], 
       aes(x = age, fill = factor(pclass))) +
    geom_density(alpha = 0.4)

```
